---
title: "Derivation of time-independent growth functions"
author: "R. Molowny-Horas"
date: "`r Sys.Date()`"
header-includes:
  - \usepackage{enumerate}
output:
  html_document: default
  pdf_document: default
  word_document: default
vignette: >
  %\VignetteIndexEntry{Derivation of time-independent growth functions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

<br> <br>

## Introduction

Growth, understood as an increment along the physical dimensions of an
organism, characterizes the temporal dynamics of many living entities
better than many other features. By physical dimensions or size, terms
that will be used indistinctly hereafter, we mean quantities that are
directly related to the spatial space occupied by an organism (e.g.
diameter and height for trees, length or weight for animals). Thus,
those physical dimensions will increase at varying rate as the organism
ages.

Size is often easier to determine experimentally than age. Only under
controlled conditions in a laboratory or field experiment can we observe
the birth or germination of animals or plants, and follow each single
organism continuously and methodically to determine exactly their life
span. However, in natural experiments like forest inventories, observers
do not have control over experimental conditions and, therefore, the age
of an individual is seldom precisely quantifiable.

Even under ideal conditions, growth of an organism is hampered by
structural and physiologically constraints, which gradually set in as
the organism gets larger. Those constraints usually make its growth rate
(i.e. increment in size per unit time) to slow down beyond some point,
approaching zero as time tends to infinite. We will hereby discard those
cases in which growth rate in size may become zero or even negative
(i.e. size reduction) due to, for example, worsening environmental
conditions or fiercer competition. It may thus seem reasonable to assume
that, for an organism, there is an upper limit for size which is only
reached at an advanced age and under the most optimum climatic and
environmental conditions. That upper limit is usually unknown but may be
guessed by extrapolating observed trends. In addition, the upper-limit
assumption suggests that sigmoid curves may be a good analytic model to
describe size as a function of time, since sigmoid curves are
monotonically increasing and are also constrained by lower and upper
asymptote.

Let $y(t)$ be a continuous and differentiable sigmoid-like curve, which
depends on time or age $t$. Given two time points $t_{1}$ and $t_{2}$,
where $t_{1}<t_{2}$, then growth , defined as $y(t_{2})-y(t_{1})$, will
simply be an unknown function $f\left(t_{1}, t_{2}\right)$. Expressed
this way, that function $f\left(t_{1}, t_{2}\right)$ depends explicitly
on absolute age/time at $t_{1}$ and $t_{2}$, which are usually unknown
quantities in natural experiments. To determine growth independent of
age/time, we will seek an analytic expression that is instead a function
of the time interval $\Delta=t_{2}-t_{1}$, a relative quantity that is
much easier to obtain. That is, we want to be find a new growth function
$g$:

```{=tex}
\begin{equation}
f\left(t_{1}, t_{2}\right)=g\left(\Delta\right)
\end{equation}
```
Arriving at such functional expression requires some algebraic
manipulation of the original, age-dependent sigmoid functions, as shown
below. Our objective, throughout this work, will be to derive
age/time-independent expressions for growth rate of the form
$g\left(\Delta\right)$ for the following six sigmoid functions:

1.  Logistic
2.  Schumacher
3.  Gompertz
4.  Monomolecular
5.  Arc tangent
6.  Hyperbolic tangent
7.  Arc tangent exponential
8.  Rational
9.  Gaussian

Those expressions should also incorporate a specific way to determine
their dependence on a given set of explanatory variables like e.g.
climate, landscape, competition, etc.

<br>

## Notation followed throughout the package

Package functions follow a notation convention that is very similar to
the one followed in the equations of this Vignette. The table below will
identify variables and their

-   $t$, $t_{1}$ and $t_{2}$: time points, where $t_{1}\leq t_{2}$.
-   $\Delta$: time difference, $\Delta = t_{2} - t_{1}$
-   $y_{1}$ and $y_{2}$: size at time points $t_{1}$ and $t_{2}$,
    respectively. It is assumed throughout that $y_{1}\leq y_{2}$.
-   $k$: growth rate parameter. The dependence of growth on a set of
    explanatory variables is conveyed through this parameter.
-   $\mathcal{X}$: set of explanatory variables on which growth will
    depend on through $k$.
-   $a_{i}$: coefficients that multiply variables in the set
    $\mathcal{X}$.
-   $A$: asymptotic maximum size.
-   $B$: offset term.
-   $g\left(\Delta\right)$: growth function such that
    $y_{2}=y_{1}+g\left(\Delta\right)$.

<br> <br>

## Methodology

We will derive time-independent (also known as age-independent) curves
for several $g\left(\Delta\right)$ functions. These curves have all been
derived from sigmoid age-dependent size curves, which implies that there
is always an asymptotic maximum size $A$. Some curves are well-known and
widely used (e.g. logistic, Schumacher, Gompertz, monomolecular) whereas
others are more rarely employed (e.g. arctangent, hyperbolic tangent).

To convey the dependence of growth on a given set of explanatory
variables $\mathcal{X}$ like e.g. climate, landscape, competition, we
will use the aforementioned parameter $k$ as a linear function:

```{=tex}
\begin{equation}
  k=\sum_{i=1}^{n} a_{i}\cdot \mathcal{X}_{i}
\end{equation}
```
which, in turn, multiplies age/time $t$. Many age-dependent size curves,
and their corresponding age-independent expressions, are highly
non-linear in $k$. Thus, a successful fitting strategy requires the
implementation of a non-linear regression algorithm to calculate the
coefficients $a_{i}$ that determine the effect of the corresponding
explanatory variable. As we will see below, this strategy yields
relatively simple expressions that can readily be implemented as
time-independent growth functions.

In addition, term $B$ represents an (a priori unknown) offset in the
growth curve. Nevertheless, as we will see below, this summative offset
$B$ conveniently vanishes in all cases when $g\left(\Delta\right)$ or
$k$ are expressed as a function of $\Delta$. Finally, in all expressions
it is easy to check that $g\left(\Delta\right)=0$ when $\Delta=0$ or
$k=0$ always, as expected.

Below we will derive the time-independent expressions for the following
six sigmoid functions:

1.  Logistic
2.  Schumacher
3.  Gompertz
4.  Monomolecular
5.  Gaussian
6.  Arc tangent
7.  Hyperbolic tangent
8.  Arctangent exponential
9.  Rational

<br>

#### 1. Logistic

The logistic curve, labelled *Autocatalytic* in Richards' original
article, is one of the most widely used sigmoid functions to determine
the growth of an individual plant or animal. In a general case, in which
size depends on time $t$ as well as on a set of predictors (e.g.
climatic, topographic), the size $y$ of an organism that obeys the
logistic rule can be written:

```{=tex}
\begin{equation}
  y=\frac{A}{1+e^{-\left(k\cdot t+B\right)}}
\end{equation}
```
As mentioned above, the dependence on predictors is done through $k$. To
obtain a time-independent equation that determines growth between two
time points $t_{1}$ and $t_{2}$, $t_{1}<t_{2}$, we start by writing:

```{=tex}
\begin{equation}
  e^{-\left(k\cdot t+B\right)}=\frac{A}{y}-1
\end{equation}
```
and:

```{=tex}
\begin{equation}
  t=\frac{1}{k}\cdot \ln\left(\frac{y}{A-y}\right)-B
\end{equation}
```
Let us define a subscript notation whereby $t_{1}$ and $t_{2}$ indicate
time points separated by $\Delta$ years, where $\Delta=t_{2}-t_{1}$.
Likewise, $y_{1}$ and $y_{2}$ will denote sizes at those time points,
respectively. Then:

```{=tex}
\begin{equation}
\Delta =t_{2}-t_{1}=\frac{1}{k}\cdot\left[\ln\left(\frac{y_{2}}{y_{1}}\cdot\frac{A-y_{1}}{A-y_{2}}\right)\right]
\end{equation}
```
The $B$ coefficient has conveniently disappeared from the equation. Now
we can easily deduce the following expression for $k$:

```{=tex}
\begin{equation}
  k=\frac{1}{\Delta}\cdot\left[\ln\left(\frac{y_{2}}{y_{1}}\cdot\frac{A-y_{1}}{A-y_{2}}\right)\right]
\end{equation}
```
Since $t_{2}=t_{1}+\Delta$, we can arrive at an expression for $y_{2}$
as a function of $y_{1}$, $\Delta$ and the coefficients $a_{i}$:

```{=tex}
\begin{equation}
  y_{2}=\frac{A}{1+e^{-\left(k\cdot t_{1}+B\right)}\cdot e^{-k\cdot\Delta}}=\frac{A}{1+\left(\frac{A}{y_{1}}-1\right)\cdot e^{-k\cdot\Delta}}
\end{equation}
```
Finally, growth can be expressed as:

```{=tex}
\begin{equation}
  g\left(\Delta\right)=\frac{A}{1+e^{-\left(k\cdot t_{1}+B\right)}\cdot e^{-k\cdot\Delta}}-y_{1}=\frac{A}{1+\left(\frac{A}{y_{1}}-1\right)\cdot e^{-k\cdot\Delta}}-y_{1}
\end{equation}
```
<br>

#### 2. Schumacher

The Schumacher growth curve can be expressed as:

```{=tex}
\begin{equation}
  y=A\cdot e^{-\frac{1}{k\cdot t+B}}
\end{equation}
```
For this curve to make ecological sense, $k\cdot t+B\geq 0$ always.
Otherwise, $y$ would grow exponentially when $k\cdot t+B< 0$. Then, we
can express $t$ as:

```{=tex}
\begin{equation}
  t=\frac{1}{k}\cdot\frac{1}{\ln\left(\frac{A}{y}\right)}-\frac{1}{k}\cdot B
\end{equation}
```
Now, following the same procedure as above, we write:

```{=tex}
\begin{equation}
  \Delta=\frac{1}{k}\cdot\left[\frac{1}{\ln\left(\frac{A}{y_{2}}\right)}-\frac{1}{\ln\left(\frac{A}{y_{1}}\right)}\right]
\end{equation}
```
As above, the offset $B$ has vanished from the equation. Then:

```{=tex}
\begin{equation}
  k=\frac{1}{\Delta}\cdot\left[\frac{1}{\ln\left(\frac{A}{y_{2}}\right)}-\frac{1}{\ln\left(\frac{A}{y_{1}}\right)}\right]
\end{equation}
```
Solving the equation for $y_{2}$:

```{=tex}
\begin{equation}
  y_{2}=A\cdot e^{-\left[\frac{1}{\frac{1}{\ln\left(\frac{A}{y_{1}}\right)}+k\cdot\Delta}\right]}
\end{equation}
```
This is exactly Equation [9] in Tomé et al. (2006). Then, the Schumacher
time-independent growth function becomes:

```{=tex}
\begin{equation}
g\left(\Delta\right)=A\cdot e^{-\left[\frac{1}{\frac{1}{\ln\left(\frac{A}{y_{1}}\right)}+k\cdot\Delta}\right]}-y_{1}
\end{equation}
```
<br>

#### 3. Gompertz

The Gompertz function has long been used by the actuarial sciences to
build life-tables. It is expressed as:

```{=tex}
\begin{equation}
  y=A\cdot e^{-e^{-\left(k\cdot t+B\right)}}
\end {equation}
```
where B is an offset. Again, we can manipulate that equation to extract
$t$:

```{=tex}
\begin{equation}
  t=\frac{1}{k}\cdot\left[\ln\left(\frac{1}{\ln\left(\frac{A}{y}\right)}\right)-B\right]
\end{equation}
```
Defining, as before, $\Delta=t_{2}-t_{1}$ we obtain:

```{=tex}
\begin{equation}
  \Delta=\frac{1}{k}\cdot\ln\left(\frac{\ln\left(\frac{A}{y_{1}}\right)}{\ln\left(\frac{A}{y_{2}}\right)}\right)
\end{equation}
```
Thus:

```{=tex}
\begin{equation}
  k=\frac{1}{\Delta}\cdot\ln\left(\frac{\ln\left(\frac{A}{y_{1}}\right)}{\ln\left(\frac{A}{y_{2}}\right)}\right)
\end{equation}
```
Finally, we arrive at:

```{=tex}
\begin{equation}
  y_{2}=A\cdot\left(\frac{y_{1}}{A}\right)^{e^{-k\cdot\Delta}}
\end{equation}
```
and:

```{=tex}
\begin{equation}
  g\left(\Delta\right)=A\cdot\left(\frac{y_{1}}{A}\right)^{e^{-k\cdot\Delta}}-y_{1}
\end{equation}
```
<br>

#### 4. Monomolecular

The monomolecular growth function does not have a smooth initial growth,
as other curves discussed in this document. It grows very fast at $t=0$
and gradually slows down until towards an asymptotic value. Its
functional expression is as follows:

```{=tex}
\begin{equation}
  y=A\cdot \left(1-e^{-\left( k\cdot t+B\right)}\right)
\end {equation}
```
If we extract $t$ we get:

```{=tex}
\begin{equation}
  t=\frac{1}{k}\cdot \left[\ln\left(\frac{A}{A-y}\right)-B\right]
\end {equation}
```
and then:

```{=tex}
\begin{equation}
  \Delta=\frac{1}{k}\cdot \left[\ln\left(\frac{A-y_{1}}{A-y_{2}}\right)\right]
\end {equation}
```
The growth rate parameter is:

```{=tex}
\begin{equation}
  k=\frac{1}{\Delta}\cdot \left[\ln\left(\frac{A-y_{1}}{A-y_{2}}\right)\right]
\end {equation}
```
Then, size at $t_{2}$ is:

```{=tex}
\begin{equation}
  y_{2}=A-\left(A-y_{1}\right)\cdot e^{-k\cdot \Delta}
\end {equation}
```
and, finally, age/time-independent growth can be written as:

```{=tex}
\begin{equation}
  g\left(\Delta\right)=\left(A-y_{1}\right)\cdot\left(1-e^{-k\cdot \Delta}\right)
\end {equation}
```
<br>

#### 5. Squared monomolecular

With a simple modification it is possible to give a smooth initial
growth to the monomolecular growth function. By squaring the term inside
the parenthesis we get:

```{=tex}
\begin{equation}
  y=A\cdot \left(1-e^{-\left( k\cdot t+B\right)^{2}}\right)
\end {equation}
```

Term $k$ is straightforward to obtain after some easy algebra:

```{=tex}
\begin{equation}
  k=\frac{1}{t}\cdot\sqrt{-\ln\left(1-\frac{y}{A}\right)}
\end {equation}
```

Then:

```{=tex}
\begin{equation}
  k\cdot\Delta=\sqrt{-\ln\left(1-\frac{y_{2}}{A}\right)}-\sqrt{-\ln\left(1-\frac{y_{1}}{A}\right)}
\end {equation}
```

Finally, we show the explicit form of $y_{2}$ as:

```{=tex}
\begin{equation}
  y_{2}=A\cdot\left(1-e^{-\left(k\cdot\Delta+\sqrt{-\ln\left(1-\frac{y_{1}}{A}\right)}\right)}\right)
\end {equation}
```

Therefore, the growth function will be:

```{=tex}
\begin{equation}
  g\left(\Delta\right)=A\cdot\left(1-e^{-\left(k\cdot\Delta+\sqrt{-\ln\left(1-\frac{y_{1}}{A}\right)}\right)}\right)-y{1}
\end {equation}
```


#### 6. Arctangent

The arctangent is another sigmoid curve from which a time-independent
growth function can be derived relatively easily. We start by expressing
growth as a function of time:

```{=tex}
\begin{equation}
  y=A\cdot \left(\frac{1}{\pi}\cdot \arctan\left(k\cdot t+B\right)+0.5\right)
\end {equation}
```
Thus, growth $y\in \left[0,A\right]$ and $B$ has been included to allow
for any possible offset in the curve, as before. We can easily get:

```{=tex}
\begin{equation}
  k\cdot t=\tan\left[\left(\frac{y}{A}-0.5\right)\cdot\pi\right]-B
\end{equation}
```
Therefore:

```{=tex}
\begin{equation}
  t=\frac{1}{k}\cdot\tan\left[\left(\frac{y}{A}-0.5\right)\cdot\pi\right]-B
\end{equation}
```
and, thus:

```{=tex}
\begin{equation}
  \Delta=\frac{1}{k}\cdot\left[\tan\left(\left(\frac{y_{2}}{A}-0.5\right)\cdot\pi\right)-\tan\left(\left(\frac{y_{1}}{A}-0.5\right)\cdot\pi\right)\right]
\end{equation}
```
Thus, the offset term $B$ has vanished from the equation. An equivalent
expression for $k$ is:

```{=tex}
\begin{equation}
  k=\frac{1}{\Delta}\cdot\left[\tan\left(\left(\frac{y_{2}}{A}-0.5\right)\cdot\pi\right)-\tan\left(\left(\frac{y_{1}}{A}-0.5\right)\cdot\pi\right)\right]
\end{equation}
```
Finally, we derive:

```{=tex}
\begin{equation}
\tan\left(\left(\frac{y_{2}}{A}-0.5\right)\cdot\pi\right)=\tan\left(\left(\frac{y_{1}}{A}-0.5\right)\cdot\pi\right)+k\cdot\Delta
\end{equation}
```
and then:

```{=tex}
\begin{equation}
y_{2}=A\cdot\left[\frac{1}{\pi}\cdot \arctan\left\{\tan\left(\left(\frac{y_{1}}{A}-0.5\right)\cdot\pi\right)+k\Delta\right\}+0.5\right]
\end{equation}
```
The growth function is, then:

```{=tex}
\begin{equation}
g\left(\Delta\right)=A\cdot\left[\frac{1}{\pi}\cdot \arctan\left\{\tan\left(\left(\frac{y_{1}}{A}-0.5\right)\cdot\pi\right)+k\Delta\right\}+0.5\right]-y_{1}
\end{equation}
```
<br>

#### 7. Hyperbolic tangent

The hyperbolic tangent growth function is seldom used in growth studies.
We start by expressing size $y$ as a function of $t$ as follows:

```{=tex}
\begin{equation}
  y\left(t\right)=A\cdot\left[\tanh\left(k\cdot t+B\right)+1\right]=
  \frac{A}{2}\left[\cdot\frac{e^{2\cdot\left(k\cdot t+B\right)}-1}{e^{2\cdot\left(k\cdot t+B\right)}+1}+1\right]
\end{equation}
```
We can then write:

```{=tex}
\begin{equation}
  y=A\cdot\frac{e^{2\cdot\left(k\cdot t+B\right)}}{e^{2\cdot\left(k\cdot t+B\right)}+1}=A\cdot\frac{1}{1+e^{-2\cdot\left(k\cdot t+B\right)}}
\end {equation}
```
This is exactly the same as the logistic function above if we substitute
$2\cdot k=k'$. Therefore, it is straightforward to arrive at the
following expressions:

```{=tex}
\begin{equation}
\Delta=\frac{1}{2\cdot k}\cdot\left[\ln\left(\frac{y_{2}}{y_{1}}\cdot\frac{A-y_{1}}{A-y_{2}}\right)\right]
\end{equation}
```
and:

```{=tex}
\begin{equation}
k=\frac{1}{2\cdot\Delta}\cdot\left[\ln\left(\frac{y_{2}}{y_{1}}\cdot\frac{A-y_{1}}{A-y_{2}}\right)\right]
\end{equation}
```
Finally, we obtain:

```{=tex}
\begin{equation}
  y_{2}=\frac{A}{1+\left(\frac{A}{y_{1}}-1\right)\cdot e^{-2\cdot k\cdot\Delta}}
\end{equation}
```
Growth can thus be expressed as:

```{=tex}
\begin{equation}
  g\left(\Delta\right)=\frac{A}{1+\left(\frac{A}{y_{1}}-1\right)\cdot e^{-2\cdot k\cdot\Delta}}-y_{1}
\end{equation}
```
Thus, the hyperbolic tangent growth function is simply a rescaled
version of the logistic growth function, o vice versa.

<br>

#### 8. Arctangent exponential

Similar to the arctangent growth function above, the arctangent
exponential follows a sigmoid shape, although the slope of its middle
part is steeper and it reaches the lower or upper plateau quicker. It
can be expressed as follows:

```{=tex}
\begin{equation}
  y=A\cdot \left(\frac{2}{\pi}\cdot \arctan\left(e^{k\cdot t+B}\right)\right)
\end {equation}
```
With some algebra we can obtain an explicit expression for the $k$:

```{=tex}
\begin{equation}
  k=\frac{1}{t}\left(\ln\left(\tan\left(\frac{\pi}{2\cdot A}\cdot y\right)\right)-B\right)
\end {equation}
```
It is also easy to get an expression for $k\cdot\Delta$, in which the
term $B$ is now gone:

```{=tex}
\begin{equation}
  k\Delta=\ln\left(\frac{\tan\left(\frac{\pi}{2\cdot A}\cdot y_{2}\right)}{\tan\left(\frac{\pi}{2\cdot A}\cdot y_{1}\right)}\right)
\end {equation}
```
Undoing this equation to express $y_{2}$ as a function of all other
terms we come to:

```{=tex}
\begin{equation}
  y_{2}=\frac{\pi}{2\cdot A}\cdot \arctan\left(\tan\left(\frac{\pi}{2\cdot A}\right)\cdot e^{k\cdot\Delta}\right)
\end {equation}
```
Finally, size growth is simply:

```{=tex}
\begin{equation}
  g\left(\Delta\right)=\frac{\pi}{2\cdot A}\cdot \arctan\left(\tan\left(\frac{\pi}{2\cdot A}\right)\cdot e^{k\cdot\Delta}\right)-y{1}
\end {equation}
```
<br>

#### 9. Rational

Rational functions can also behave in a sigmoid-like fashion. In
particular, the following function:

```{=tex}
\begin{equation}
y=\frac{A}{2}\cdot\left(\frac{k\cdot t+B}{\sqrt{1+\left(k\cdot t+B\right)^{2}}}+1\right)
\end {equation}
```
It is straightforward to derive an explicit expression for $k$:

```{=tex}
\begin{equation}

k=\frac{1}{t}\cdot \frac{1}{\sqrt{\frac{1}{2\cdot\frac{y}{A}-1}}}-B
\end {equation}
```
As before, evaluating this equation in $t_{1}$ and $t_{2}$ enables us to
show that:

```{=tex}
\begin{equation}
k\cdot\Delta = \frac{y_{2}}{\sqrt{1-y_{2}^{2}}} - \frac{y_{1}}{\sqrt{1-y_{1}^{2}}}
\end {equation}
```
Reordering and squaring we get:

```{=tex}
\begin{equation}
\frac{y_{2}^{2}}{1-y_{2}^{2}}=\frac{1}{\frac{1}{y_{2}^{2}}-1}=\left(k\cdot\Delta + \frac{y_{1}}{\sqrt{1-y_{1}^{2}}}\right)^{2}
\end {equation}
```
Finally, some further algebraic manipulation allows to write:

```{=tex}
\begin{equation}
y_{2}=\sqrt{\frac{1}{1+\frac{1}{\left(k\cdot\Delta + \frac{y_{1}}{\sqrt{1-y_{1}^{2}}}\right)^{2}}}}
\end {equation}
```
Thus, growth in the case of this rational function will be given by:

```{=tex}
\begin{equation}
g\left(\Delta\right)=\frac{A}{\sqrt{1+\frac{1}{\left(k\cdot\Delta + \frac{y_{1}}{\sqrt{1-y_{1}^{2}}}\right)^{2}}}}-y_{1}
\end {equation}
```
<br>

### Comparison of curves.

Below we show an example plot for five time-dependent growth functions
available in this package. The hyperbolic tangent curve is missing since
it is identical to a rescaled logistic. The schumacher, monomolecular and squared monomolecular curves are
not well defined for points $k\cdot t_{min}+B\lt 0$. For that reason
those points to the left in those curves are set to zero in the plot.

```{r fig.align="center", width=8}
t <- seq(0, 200, by = .1) ; k  <- 0.1 ; B <- -2
dat <- data.frame(t = t, k = 0.1, offset = -2, max_y = 120)
df <- data.frame(x = t, y = Rtigre::td_size(dat, curve_type = "logistic"), curve_type = "logistic")
for (i in Rtigre:::all_curve_types()[-c(1, 10)]) {
  y <- Rtigre::td_size(dat, curve_type = i)
  if (any(i %in% c("schumacher", "monomolecular", "monomolecular2"))) y[(k*t+B) < 0] <- 0
  df <- rbind(df, data.frame(x = t, y = y, curve_type = i))
}
print(ggplot2::ggplot(df, ggplot2::aes(x = x, y = y, colour = curve_type)) + ggplot2::geom_line())
```

<br>

The figure below shows the time-independent versions of the same curves. The time-independent monomolecular curve is monotonically decreasing and only its lower part is shown. Color code is the same that has been followed in the previous figure above.

```{r fig.align="center", width=8}
tdiff <- 1
max_y <- 120
y1 <- seq(0, max_y, length = 1000)
k <- .1
dat <- data.frame(tdiff = tdiff, max_y = max_y, k = k, y1  = y1)
plot(y1, Rtigre::ti_size(dat)-dat$y1, xlab = "Size at t1",
     ylab = "Size at t2 - size at t1", type = "l", lwd = 2, col = 1, ylim = c(0, 10))
points(y1,Rtigre::ti_size(dat, "schumacher")-dat$y1, type = "l", lwd = 2, col = 2)
points(y1,Rtigre::ti_size(dat, "gompertz")-dat$y1, type = "l", lwd = 2, col = 3)
points(y1,Rtigre::ti_size(dat, "monomolecular")-dat$y1, type = "l", lwd = 2, col = 4)
points(y1,Rtigre::ti_size(dat, "monomolecular2")-dat$y1, type = "l", lwd = 2, col = 5)
points(y1,Rtigre::ti_size(dat, "arctangent")-dat$y1, type = "l", lwd = 2, col = 6)
points(y1,Rtigre::ti_size(dat, "rational")-dat$y1, type = "l", lwd = 2, col = 7)
points(y1,Rtigre::ti_size(dat, "arctangent_exp")-dat$y1, type = "l", lwd = 2, col = 8)
```
