---
title: "Explanatory factor variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explanatory_factor_variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

Unlike other R functions like e.g. "lm" or "glm", syntax requirements for formulas in function "nls" does not accept factors as such. The


## Example data

To show how this all works we will use the "iris" dataset.


```{r fig.width = 5,  fig.align = "center"}
ggplot2::ggplot(iris, ggplot2::aes(x = Sepal.Length, y = Sepal.Width,
               group = Species, colour = Species)) + 
    ggplot2::geom_point() +
    ggplot2::scale_colour_manual(values = c("red", "blue", "green")) +
  ggplot2::labs(x = "Sepal length", y = "Sepal width")
```

<br>

The "Species" column in the "iris" data.frame is coded as a factor with 3 levels containing the name of the 3 species of plants. For a successful non-linear regression scheme with the built-in "nls" function or the "nlsLM" from the "minpack.lm" R package we must code the factor variable as several *dummy* variables. In our example with the "iris" dataset, if the factor "Species" has 3 levels we must build 3 dummy variables containing 0's and 1's.

```{r}
iris$setosa <- iris$Species == "setosa"
iris$versicolor <- iris$Species == "versicolor"
iris$virginica <- iris$Species == "virginica"
```

<br>

Although a linear model would probably be good enough for the "iris" dataset, we will apply a non-linear model to demonstrate how one can implement it in "nls". In our case, we will assume that sepal width is a function of sepal length of the form:

```{=tex}
\begin{equation}
f\left(y\right)=c1\cdot\ e^{-\left(c2\cdot x+Species\right)}
\end{equation}
```

where, for the sake of conciseness, "x" stands for sepal length and "y" for sepal width.
<br>

Since function "nls" does not accept factor "Species" as a valid explanatory variable, we will use the dummy variables calculated above.

```{r}
m1 <- nls(Sepal.Width ~ c1 * exp(-(c2*Sepal.Length + c3*setosa + c4*versicolor)), iris,
         start = list(c1 = 1, c2 = .01, c3 = 0, c4 = 0))

m2 <- nls(Sepal.Width ~ exp(-(c2*Sepal.Length + c3*setosa + c4*versicolor + c5*virginica)), iris,
         start = list(c2 = .01, c3 = 0, c4 = 0,  c5 = 0))
```

<br>

Notice that $c1=e^{-c5}$, as expected.

In this example we have introduced first guesses (the "start" list) that are sufficiently close to the final solution that the algorithm in "nls" converges quickly and smoothly. Those values were produced through a trial-and-error hand procedure in which the "nls" function was run with several different parameters until successful convergence could be reached. However, when the non-linear expression to be fitted is very complex, it is usually very hard or almost impossible to guess start values for the regression parameters.

Several strategies can be implemented to find reasonable start values for a non-linear problem. For example, if an approximate range of values for each parameter is known, it should be possible to automatize the trial-and-error strategy mentioned above until convergence is guaranteed. Another possibility would be to treat the regression as a minimization problem and use the "optim" function with the robust, though slow, "Nelder-Mean" method, which requires less precise starting values to converge.

In the case of the highly non-linear growth equations exemplified in this package, those procedures are probably too hard to implement and have no guarantee of succeeding. However, as discussed in an accompanying vignette, we can algebraically derive alternative expressions that can then be used to obtain first guesses for the parameters easily.



## Growth equations

We will now turn to a "real" case and will apply the growth equations implemented in this package.

```{r}
library(Rtigre)
data("Punci_IFN")
Punci_IFN$Treatment <- sample(LETTERS[1:3], nrow(Punci_IFN), replace = T)
Punci_IFN$tdiff <- 10
```


<br>

Next, we codify the 3-level factor "Treatment" in as many new dummy variables (although we will use only two of them).

```{r}
Punci_IFN$A <- (Punci_IFN$Treatment == "A")*1
Punci_IFN$B <- (Punci_IFN$Treatment == "B")*1
Punci_IFN$C <- (Punci_IFN$Treatment == "C")*1
r <- fit_growth(Punci_IFN, ~ prec + temp + A + B, verbose = F)
```

